### 2020-05-19 manuscript.Rmd

@@

Insufficient physical activity is a leading cause of disease and disability worldwide.[@who2009] For the UK, around one in six deaths is attributable to low levels of physical activity.[@lee2012] It is also a major contributor to health inequalities, as people from low socio-economic backgrounds are both disproportionately likely to be inactive [@rowe2000sport; @farrell2014] and be affected by physical inactivity-related diseases. Increasing the physical activity levels of the population is therefore high on the public health agenda: it not only has the potential to improve quality of life, reduce mortality rates and alleviate the strain on health and social care services, but also reduce the gap in health inequalities [@vilhelmsson2018].

@@

Notwithstanding these subjective accounts, the expansion of parkrun in England, as elsewhere, has been largely driven by *demand*, not *need*. It might therefore be the case that parkrun events are primarily located in areas that are relatively well-off anyway, while people living in more deprived communities may not have the same opportunities to participate. In 2019, Sport England announced funding to support the creation of 200 new parkrun events across England within three years, with the specific aim of increasing participation of individuals from lower socio-economic groups.[@reuters2018]

The aims of this study are two-fold: firstly, to evaluate whether geographic access to parkrun events in England geographic access to parkrun events in England is equitable across areas with different levels of deprivation. Secondly, to identify 200 optimal locations for future events in order geographic access to parkrun events in England is equitable across areas with different levels of deprivation. Secondly, to identify 200 optimal locations for future events in order to improve geographic access, in particular for deprived communities.

@@

This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units which, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socio-economic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on public green spaces in England to conduct a simple location-allocation analysis, in order to identify 200 locations for future parkrun events, that maximise access for the population.

@@

  3) The locations of public green spaces in England were retrieved from an open dataset of Ordnance Survey.[@ordnance] Even though parkrun events do not always require a 5km loop (some courses involve running back and forth) and are held in various settings and terrains, not all green spaces are suitable to host running events. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields with an area of 0.1 km$^2$ (e.g. 316m x 316m) or more potentially suitable for hosting events (n=2,842).

@@

### Association between Deprivation and Access

There was a negative relationship between IMD and the distance to the nearest parkrun event: the (population weighted) Pearson and Spearman correlation coefficients were -0.15 and -0.18, indicating a small effect. This means, more deprived LSOAs tended to have shorter distances to the nearest parkrun event, i.e. better access.

@@

We estimated that setting up new parkrun events in those 200 green spaces would improve access for around 16.5 million people (30% of the population) from 9,854 LSOAs. For those, the distance to the nearest event would, on average, be reduced by 4.09 km (SD = 3.97). Overall, it would reduce the average and median distance to the nearest parkrun event from 4.65 and 3.39 km to 3.43 and 2.59 km. The percentage of people who live within 5km of a parkrun would increase from 69% to 81.9%. Figure 2 shows the distribution of distances between LSOAs and their nearest parkrun events before and after the introduction of the 200 new optimal event locations at the identified optimal green spaces.

@@

  Overall, geographic access to parkrun in England is good: the median distance to the nearest parkrun event was 3.39 km and more than two thirds of the population lived within 5km (the parkrun distance) of a parkrun event. Contrary to our expectation, we did not find that access was better for people living in more affluent areas. In fact, those living in the most deprived areas had the best geographic access to parkrun. It is rare in public health for inequalities to exist in this direction.

@@

  The main finding, that geographical access to parkrun events is better in more deprived communities, is surprising. parkrun events are set up by volunteers, based on demand, not need. Studies have shown that the level of physical activity, and the availability of physical activity facilities generally declines with the level of deprivation. Opportunities for physical activity are often lacking in areas in most need.[e.g. @hillsdon2007] parkrun events, in contrast, seem to be often held in or near deprived areas, and are free to attend, giving anyone equal access, irrespective of their socio-economic background. Nevertheless, in a previous analysis, we found that participation in parkrun has a strong socioeconomic gradient with considerably higher participation rates in more affluent areas: about a third of all participants came from LSOAs in the least deprived quintile, while only 7% came from the most deprived quintile.[@smithschneider; @schneidersmith] This suggests that providing the opportunity to participate in parkrun events, while a necessary first step to enable participation, has not been sufficient to engage deprived communities.[@white2009]

@@

  In England, geographic access to parkrun events is good, and can be further improved with the creation of 200 new events. Contrary to our expectation, we find that geographic access is better for those living in more deprived communities. Given that participation rates are generally lower in deprived areas, improving access alone is unlikely to significantly reduce inequalities in participation and/or levels of physical activity. A deeper understanding is needed of the barriers to participation in deprived communities if strategies are going to be created to improve engagement.

>>>>

Corresponding author: Paul Schneider (p.schneider@sheffield.ac.uk)

Target Journal: 1) Public Health (https://www.journals.elsevier.com/public-health), 2) BMJ Open, 3) Wellcome Open Research

## Abstract

### Objectives

To evaluate geographical access to parkrun in England, and to identify 200 locations for future events to maximize geographic access to parkrun.

This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018.

### Methods

We combined geo-location data on all English Lower Layer Super Output Areas and parkrun events to calculate geodesic distances to the nearest event for over 32,000 communities in England. We use this measure of geographic access to summarise the relationship between access and socio-economic deprivation, measured using the Index of Multiple Deprivation. We then used geographic coordinates of public green spaces in England to conduct a simple location-allocation analysis to identify 200 locations for future parkrun events that would maximize access.

### Results

Geographical access to parkrun in England is good, over two thirds of the population live within 5km of a parkrun. It is also very equitable, access is slightly better in more deprived communities. Optimally locating an additional 200 events would improve access considerably, approximately 82% of the English population would be within 5km of a parkrun event.

### Conclusions

Access to parkrun is good, and equitable. Creating new parkrun events will improve geographic access to parkrun, but a deeper understanding of the barriers to participation in deprived communities is necessary in order to develop effective strategies to improve engagement.

@@

Insufficient physical activity is one of the leading causes of disease and disability worldwide.[@who2009] In the UK, around one in six deaths is attributable to low levels of physical activity.[@lee2012] It is also a major contributor to health inequalities, as people from low socio-economic backgrounds are both disproportionately likely to be inactive [@rowe2000sport; @farrell2014] and be affected by physical inactivity-related diseases [@petrovic2018contribution]. Increasing the physical activity levels of the population is therefore high on the public health agenda: it not only has the potential to improve quality of life, reduce mortality rates and alleviate the strain on health and social care services, but also reduce the gap in health inequalities [@vilhelmsson2018].

@@

Notwithstanding these subjective accounts, the expansion of parkrun in England, as elsewhere, has been largely grassroots, driven by *demand* rather than *need*. It might therefore be the case that parkrun events are primarily located in areas that are relatively affluent, while people living in more deprived communities may not have the same opportunities to participate. In 2019, Sport England announced funding to support the creation of 200 new parkrun events across England within three years, with the specific aim of increasing participation of individuals from lower socio-economic groups.[@reuters2018]

The aims of this study are two-fold: firstly, to evaluate whether geographic access to parkrun events in England geographic access to parkrun events in England is equitable across areas with different levels of deprivation. Secondly, to identify 200 optimal locations for future events in order to improve geographic access, in particular for deprived communities.

@@

This study is a cross-sectional ecological analysis of the socio-economic disparities in geographic access to parkrun events in England at the end of 2018. All analyses were conducted on the level of Lower Layer Super Output Areas (LSOAs), which divide England into 32,844 geographic units which, on average, have a population of approximately 1,700. We assessed the relationship between access, defined as the distance (as the crow flies) to the nearest parkrun event, and socio-economic deprivation, measured using the Index of Multiple Deprivation. In addition, we used information on public green spaces in England to conduct a simple location-allocation analysis, in order to identify 200 locations for future parkrun events that maximise access for the population.

@@

  3) The locations of public green spaces in England were retrieved from an open dataset of Ordnance Survey.[@ordnance] parkrun events are held in various settings and terrains, and do not always require a single 5km loop - some events have courses that involve running back and forth. After evaluating existing parkrun event courses, we decided to consider all public parks, gardens, and playing fields in England with an area of 0.1 km$^2$ or more potentially suitable for hosting events (n=2,842).

@@

### Association between deprivation and access

There was a negative relationship between IMD and the distance to the nearest parkrun event: the (population weighted) Pearson and Spearman correlation coefficients were -0.15 and -0.18, indicating a small negative correlation. This means that more deprived LSOAs tended to have shorter distances to the nearest parkrun event, i.e. better geographical access.

@@

We estimated that setting up new parkrun events in those 200 green spaces would improve access for around 16.5 million people (30% of the population) from 9,854 LSOAs. For these people, the distance to the nearest event would, on average, be reduced by 4.09 km (SD = 3.97). Overall, it would reduce the average and median distance to the nearest parkrun event from 4.65 and 3.39 km to 3.43 and 2.59 km. The percentage of people who live within 5km of a parkrun would increase from 69% to 82%. Figure 2 shows the distribution of distances between LSOAs and their nearest parkrun events before and after the introduction of the 200 new optimal event locations at the identified optimal green spaces.

@@

  Overall, geographic access to parkrun in England is good: in December 2018 the median distance to the nearest parkrun event was 3.39 km and more than two thirds of the population lived within 5km (the parkrun distance) of a parkrun event. Contrary to our expectation, we did not find that access was better for people living in more affluent areas. In fact, those living in the most deprived areas had the best geographic access to parkrun. It is rare in public health for inequalities to exist in this direction.

@@

  The main finding, that geographical access to parkrun events is better in more deprived communities, is surprising. parkrun events are set up by volunteers, based on demand, not need. Studies have shown that the level of physical activity, and the availability of physical activity facilities generally declines with the level of deprivation.[@farrell2014socioeconomic] Opportunities for physical activity are often lacking in areas in most need.[e.g. @hillsdon2007] parkrun events, in contrast, seem to be often held in or near deprived areas, and are free to attend, giving anyone equal access, irrespective of their socio-economic background. Nevertheless, in a previous analysis, we found that participation in parkrun has a strong socioeconomic gradient with considerably higher participation rates in more affluent areas: about a third of all participants came from LSOAs in the least deprived quintile, while only 7% came from the most deprived quintile.[@smithschneider; @schneidersmith] This suggests that providing the opportunity to participate in parkrun events, while a necessary first step to enable participation, has not been sufficient to engage deprived communities.[@white2009]

@@

  In England, geographic access to parkrun events is good, and can be further improved with the creation of 200 new events. Contrary to our expectation, we find that geographic access is better for those living in more deprived communities. Given that participation rates are generally lower in deprived areas, improving access alone is unlikely to significantly reduce inequalities in participation and/or levels of physical activity. A deeper understanding is needed of the barriers to participation in deprived communities if effective strategies are going to be created to improve engagement.

## Additional Information

### Grant Information
P.S, R.S & T.B are joint funded by the Wellcome Trust Doctoral Training Centre in Public Health Economics and Decision Science [108903] and the University of Sheffield.

### Competing Interests
P.S, R.S, T.B & A.Br have no competing interests.
S.H. is chair, A.Bu. and H.Q are deputy chairs, and E.G. is a member of the parkrun research board.

### Software & Data Availability
All data and source code are available at: TBC (<add final GitHub repo>).

### Author contributions

Paul Schneider; Roles: Conceptualization, Data Curation, Formal Analysis, Investigation, Methodology, Project Administration, Software, Visualization, Writing – Original Draft Preparation, Writing – Review & Editing

Robert Smith; Roles: Conceptualization, Data Curation, Formal Analysis, Investigation, Methodology, Project Administration, Software, Visualization, Writing – Original Draft Preparation, Writing – Review & Editing

Alice Bullas; Roles: Writing – Review & Editing

Steve Haake; Writing – Review & Editing

Helen Quirk; Roles: Writing – Review & Editing

Thomas Bayley; Roles:  Methodology

Alan Brennan; Roles: Supervision, Methodology, Writing – Review & Editing

Elizabeth Goyder; Roles: Supervision, Conceptualization, Writing – Review & Editing

## References
